---
title: "sysVar_inOut_V05"
author: "Emily Butler, eabutler@u.arizona.edu & Ashley Kuelz, akuelz@email.arizona.edu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sysVar_inOut_V05}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Once you have gone through the steps for estimating dynamic trajectory profiles for one of the models (see "inertia_coordination" and "coupled_oscillator" vignettes), you are ready to use the profiles as either predictors or outcomes of system variables, e.g., any variables of interest that you think are related to your dyadic system, but change slower than the state variables used to represent the dynamics. We show only the basic steps for getting the profile memberships from the inertia-coordination model here (see other vignettes for full procedure).

```{r, results= "hide"}
library(rties)

data1 <- rties_ExampleData_4

data2 <- dataPrep(basedata=data1, dyadId="couple", personId="person", obs_name="dial", dist_name="female", time_name="time", time_lag="absMaxCC") 

ic <- indivInertCoord(prepData=data2, whichModel="inertCoord")

lpaData <- inspectProfiles(whichModel="inertCoord", prepData=data2, paramEst=ic$params, n_profiles=2) 

fullData <- makeFullData(basedata=data1, dyadId="couple", personId="person", dist_name="female", lpaData=lpaData, params=ic$params)
```


### Predicting the System Variable From the Profiles 

The "sysVarOut" function uses the profile memberships to predict system variables, which can be either dyadic (sysVarType = "dyadic"), where both partners have the same score (e.g., relationship length) or individual (sysVarType = "indiv"), where the partners can have different scores (e.g., age). It takes as arguments the name of the dataframe containing the profile membership scores combined with your original dataframe (created by the "makeFullData" function), the name of the column in the dataframe containing the variable you would like to use as the system variable, and whether the system variable is "dyadic" or "individual". For dyadic system variables, the only predictor is profile membership and the model is a regular regression model since all variables are at the level of the dyad. If the system variable is individual then the model is a random-intercept dyadic model and 3 models are estimated: 1) the main effect of profile membership ("profile"), 2) main effects of profile membership and the distinguishing variable ("profilePlusDist"), and 3) the interaction of profile membership and the distinguishing variable ("profileByDist"). If the system variable is not normally distributed, any of the generalized linear models supported by glm (for dyadic system variables) or glmmPQL (for individual system variables) are available by specifying the "family" distribution (use ?sysVarOut for more information).

For normally distributed system variables, the function returns a list including the lm or lme objects containing the full results for each model (called "models"). These results can be inspected using the usual "summary" function and the models can be compared using the "anova" function. Similarly, for non-normal system variables, the function returns a list of the glm or glmmPQL objects containing the full results for the models. The function also displays a boxplot of the system variable for each profile, histograms of the residuals and plots of the predicted values against observed values for each model. In this example, we find a main effect of sex, but no effect of profile (see summary and anova output), nor an interaction. The boxplot helps to interpret this result and shows that the system variable (selfsatPhys, which is self-reported satisfaction with one's own physical shape) is predicted to be higher for men than women, especially in Profiles 1 & 2.

```{r}
sysOut <- sysVarOut(fullData=fullData, sysVar_name="conflict", sysVarType="indiv", dist0name="men", dist1name="women")
```

### Predicting Profile Membership From the System Variable

Finally, you can turn the direction of prediction around and use the system variable to predict couples' profile membership. The function "sysVarIn" accomplishes this. It takes as arguments the name of the dataframe containing the profile membership scores combined with your original dataframe (created by the "makeFullData" function), the name of the column in the dataframe containing the variable you would like to use as the system variable, whether the system variable is "dyadic" or "individual", the number of profiles, and optional names to use in the plots for the levels of the distinguisher and the system variable. If there are 2 profiles, then binomial regression models are used. If there are more than 2 profiles then multinomial regression is used. For dyadic system variables, a couple's shared score is the only predictor of their profile membership (the model is called "sysVarMain"). For individual system variables, two models are tested, one with the main effects of both partner's system variable ("sysVarMain") and one with the main effects and their interaction ("sysVarInteract"). In both cases an intercept-only model is included as a comparison point (called "base").

The function returns a list of the full model results and produces plots of profile membership against the system variable(s). Due to these being either logistic or multinomial models, interpretation is somewhat more complicated than for the sysVarOut results. In this example, it is a multinomial model since we used 3 profiles. To aid intrepretation, we exponentiate the coefficients from the sysVarInteract model to get odds ratios and then calculate p-values for them. The results show evidence of an interaction effect, such that male and female self-satisfaction with physique predict membership in Profile 3 differently than they do for Profile 1. The accompanying plots show that Profile 1 is most likely when both partners are low on selfsatPhys, while Profile 3 is most likely for couples where the woman is high on selfsatPhys ane the male is low.
```{r, fig.width=5, fig.height=5}
sysIn <- sysVarIn(fullData=fullData, sysVar_name="selfsatPhys", sysVarType="indiv", n_profiles=3,
                  dist0name="men", dist1name="women")
summary(sysIn$models$sysVarInteract)
exp(coef(sysIn$models$sysVarInteract))
z <- summary(sysIn$models$sysVarInteract)$coefficients/summary(sysIn$models$sysVarInteract)$standard.errors
p <- (1 - pnorm(abs(z), 0, 1)) * 2
round(p, 3)
```

### Predicting the System Variable From the Profiles 

The next step in the analysis is to use each dyad's profile membership to predict the system variable using the "sysVarOut" function. The system variable can be either dyadic (sysVarType = "dyadic"), where both partners have the same score (e.g., relationship length) or individual (sysVarType = "indiv"), where the partners can have different scores (e.g., age). It takes as arguments the name of the dataframe containing the profile membership scores combined with your original dataframe (created by the "makeFullData" function), the name of the column in the dataframe containing the variable you would like to use as the system variable, and whether the system variable is "dyadic" or "individual". For dyadic system variables, the only predictor is profile membership and the model is a regular regression model since all variables are at the level of the dyad. If the system variable is individual then the model is a random-intercept dyadic model and 3 models are estimated: 1) the main effect of profile membership ("profile"), 2) main effects of profile membership and the distinguishing variable ("profilePlusDist"), and 3) the interaction of profile membership and the distinguishing variable ("profileByDist"). If the system variable is not normally distributed, any of the generalized linear models supported by glm (for dyadic system variables) or glmmPQL (for individual system variables) are available by specifying the "family" distribution (use ?sysVarOut for more information).

For normally distributed system variables, the function returns a list including the lm or lme objects containing the full results for each model (called "models"). These results can be inspected using the usual "summary" function and the models can be compared using the "anova" function. Similarly, for non-normal system variables, the function returns a list of the glm or glmmPQL objects containing the full results for the models. For normally distributed variables, the function also displays a boxplot of the system variable for each profile, histograms of the residuals and plots of the predicted values against observed values for each model. In this example, we predict a binary variable that indicates whether a person scored high or low on a self-reported relationship ambivalence scale. The results show a main effect of profile, such that couples in Profile 2 are more likely to report high ambivalence. To aid in interpretation, since this is a logistic regression, we can use the sjPlot and sjmisc packages to produce a plot of the predicted probabilities of high ambivalence for men and women in each profile.
```{r, fig.width=4}
sysOut <- sysVarOut(fullData=fullData, sysVar_name="ambivB", sysVarType="indiv", dist0name="Men",
                    dist1name="Women", family="binomial")
summary(sysOut$models$profileByDist)
summary(sysOut$models$profilePlusDist)
summary(sysOut$models$profile)

library(sjPlot)
library(sjmisc)
plot_model(sysOut$models$profileByDist, type="pred", terms=c("profile", "dist"))
```

### Predicting Profile Membership From the System Variable

Finally, you can turn the direction of prediction around and use the system variable to predict couples' profile membership. The function "sysVarIn" accomplishes this. It takes as arguments the name of the dataframe containing the profile membership scores combined with your original dataframe (created by the "makeFullData" function), the name of the column in the dataframe containing the variable you would like to use as the system variable, whether the system variable is "dyadic" or "individual", the number of profiles, and optional names to use in the plots for the levels of the distinguisher and the system variable. If there are 2 profiles, then binomial regression models are used. If there are more than 2 profiles then multinomial regression is used. For dyadic system variables, a couple's shared score is the only predictor of their profile membership (the model is called "sysVarMain"). For individual system variables, two models are tested, one with the main effects of both partner's system variable ("sysVarMain") and one with the main effects and their interaction ("sysVarInteract"). In both cases an intercept-only model is included as a comparison point (called "base").

The function returns a list of the full model results and produces plots of profile membership against the system variable(s). Due to these being either logistic or multinomial models, interpretation is somewhat more complicated than for the sysVarOut results. In this example, it is a logistic model since we used 2 profiles and there is no evidence of any effects, although the effect of women's ambivalence on the probability of being in Profile 2 is marginal (p < .07). The plot shows the results, or we can exponentiate the coefficients to obtain odds ratios, which shows that ambivalent women are 2.4 times more likely to be in Profile 2 than Profile 1, compared to non-ambivalent women.
```{r, fig.width=4}
sysIn <- sysVarIn(fullData=fullData, sysVar_name="ambiv", sysVarType="indiv", n_profiles=2, 
                  dist0name="Men", dist1name="Women")
summary(sysIn$models$sysVarInteract)
summary(sysIn$models$sysVarMain)
exp(coef(sysIn$models$sysVarMain))
```













